Vigilantism and Institutions:\\ Understanding Attitudes toward Lynching in Brazil

### 2023-01-03 article/article.tex
production version
small changes to intro

% \noindent \textbf{Keywords:} Brazil, crime, extralegal violence, lynching, vigilantism
\vspace{.25cm}

@@

The regulation of violence is a crucial role of the state. When states control
violence effectively, markets, politics, and civil society can flourish
\citep{besley2011pillars, north2009violence}. However, many states in the
Global South still struggle to maintain a monopoly on the use of force.
Scholars have written extensively about how rebel organizations build support
to challenge government institutions, yet we know little about what motivates
local extrajudicial violence and vigilantism in the developing world
\citep{bateson2020politics}.\footnote{On vigilantism more generally, see
\citet{cohen2022collective}, \citet{schuberth2013challenging},
\citet{smith2019contradictions}, and \citet{zizumbo2017community}.} This
omission is significant not only because vigilantism can deepen group enmities
and lead to cycles of violence, but also because it undermines the legitimacy
of the state and the rule of law, which are essential for economic development
and democratic stability \citep{jung2020lynching, tankebe2009self}.

One of the most serious forms of extrajudicial violence is lynching. Lynching
can be defined as ``incidents of physical violence committed by large numbers
of private citizens against one or more individuals accused of having committed
a ``criminal'' offense, whether or not this violence resulted in the death of
the victim(s)'' \citep[645]{godoy2004justice}. Although lynchings occur in more
than one hundred countries in all regions of the world
\citep{jung2020lynching,smith2019contradictions}, Latin America has been
particularly affected by a sharp increase in vigilante violence. Lynching
episodes have been reported in Guatemala, Mexico, Peru, Venezuela and other
countries in the region \citep{barbara2015vigilantes, cruz2019determinants,
godoy2004justice}. From 2011 to 2015, Brazil registered about 2,500 lynching
episodes, and 173 people were killed by angry mobs in 2015 alone---nearly one
execution every two days \citep{barbara2015vigilantes, oliveira2016mob}.
According to José de Souza Martins \citeyearpar{martins2015linchamentos}, who
has studied lynchings in Brazil for more than thirty years, these figures are
not only the highest in the country's history, but among the highest in the
world. The people who participate in lynchings are typically young men, but
they also sometimes include teenage women and girls, elderly women, and even
members of the local police \citep{moura2017linchamentos}.

>>>>

\noindent \textbf{Keywords:} Brazil, crime, extralegal violence, lynching, vigilantism
\vspace{.25cm}

@@

The regulation of violence is a crucial role of the state. When states control
violence effectively, markets, politics, and civil society can flourish
\citep{besley2011pillars, north2009violence}. However, many states in the
Global South still struggle to maintain a monopoly on the use of force.
Scholars have written extensively about how rebel organizations challenge
government institutions, yet we know little about what motivates local
extrajudicial violence and vigilantism in the developing world
\citep{bateson2020politics}.\footnote{On vigilantism more generally, see
\citet{cohen2022collective}, \citet{schuberth2013challenging},
\citet{smith2019contradictions}, and \citet{zizumbo2017community}.} This
omission is significant not only because vigilantism can deepen group enmities
and lead to vicious cycles of violence, but also because it undermines the rule
of law and the legitimacy of the state, which are essential for long-term
economic development and democratic stability \citep{jung2020lynching,
tankebe2009self}.

One of the most serious forms of extrajudicial violence is lynching. Lynching
can be defined as ``incidents of physical violence committed by large numbers
of private citizens against one or more individuals accused of having committed
a ``criminal'' offense, whether or not this violence resulted in the death of
the victim(s)'' \citep[645]{godoy2004justice}. Although lynchings occur in more
than one hundred countries in all regions of the world
\citep{jung2020lynching,smith2019contradictions}, Latin America has been
particularly affected by a sharp increase in vigilante violence. Lynching
episodes have been reported in Guatemala, Mexico, Peru, Venezuela, and many
other countries in the region \citep{barbara2015vigilantes,
cruz2019determinants, godoy2004justice}. From 2011 to 2015, Brazil registered
about 2,500 lynching episodes, and 173 people were killed by angry mobs in 2015
alone---nearly one execution every two days \citep{barbara2015vigilantes,
oliveira2016mob}. According to José de Souza Martins
\citeyearpar{martins2015linchamentos}, who has studied lynchings in Brazil for
more than thirty years, these numbers are not only the highest in the country's
history, but among the highest in the world. The people who participate in
lynchings are typically young men, but they also sometimes include teenage
girls, elderly women, and even members of the local police
\citep{moura2017linchamentos}.
